---
title: "Transcriptomic level adaptive capacity play roles in the deterministic assembly of a two species cocultures"
author: "Chun-Hui Gao, Hui Cao, Peng Cai"
date: "2020-11-16"
output: bookdown::html_document2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width = 8,fig.asp=0.618,message = F,dpi = 300)

# 作图
library(tidyverse)
library(ggplot2)
library(cowplot)
library(ggpubr)
library(corrplot)
library(pheatmap)
library(RColorBrewer)
library(vegan)
library(DESeq2)
library(openxlsx)
library(clusterProfiler)
theme_set(theme_bw())

# load user defined functions
source("./functions/functions.R")
source("./functions/ggplot_smooth_func.R")
 
```

## Growth of *E. coli* and *P. putida* in monoculture and cocultures

`qPCR_data` contains the result of species-specific quantitative PCR.
It has five columns:

- sample: Sample name
- condition: monoculture and coculture names. The two-species coculture system has five different initial abundance, two of which were monocultures, and the other three had the initial ratio (EC/PP) of 1:1000, 1:1, 1000:1, respectively. For convenience, we named the *P. putida* monoculture, 1:1000, 1:1, 1000:1 cocultures and *E. coli* monoculture as related to the proportion of *E. coli* to five groups of “none”, “less”, “equal”, “more” and “all”, respectively. 
- time: sample time.
- species: species name. In cocultures, the quantity was related to two species, respectively.
- quantity: the quantity of species, as calculated from the qPCR CT value and standard curves. 


```{r qPCR-data}
qPCR_data <- read.csv("data/qPCR-data.csv")
qPCR_data$condition <- factor(qPCR_data$condition, 
                              levels = c("none","less","equal","more","all"))

head(qPCR_data)

organism <- c("EC","PP")
organsim_fullname <- c("EC"="E. coli","PP"="P. putida")
```
Figure \@ref(fig:growth-curve) shows Growth curves of *E. coli* and *P. putida* in monoculture (A) and the “1:1000”, “1:1”, “1000:1” cocultures (B-D). The quantities were determined using species-specific qPCR as described in methods. In B-D subplots, the growth curves of monocultures were placed on the background layer (dashed lines), showing the difference between monoculture and cocultures.


```{r growth-curve, fig.asp=0.8, fig.cap="Growth curves of *E. coli* and *P. putida* in monoculture (A) and the 1:1000, 1:1, 1000:1 cocultures (B-D)"}

# stats
growth_data <- qPCR_data %>% group_by(species,time,condition)  %>%
  summarise(y=median(log10(quantity)),std=sd(log10(quantity),na.rm = T))

# separate monoculture and coculture
monoculture <-  growth_data %>%
  filter(condition %in% c("none","all"))
coculture <- growth_data %>%
  filter(condition %in% c("less","equal","more"))

# plot
growth_curve_mono <- ggplot(monoculture,aes(time,y,color=species)) +
  geom_line(lty="dashed",size=1) +
    geom_errorbar(aes(ymin=y-std,ymax=y+std),size=.5) + 
    geom_point(data = monoculture) 
growth_curves_coculture <- lapply(c("less","equal","more"), function(x){
  ggplot(mapping = aes(time,y,color=species)) + 
    geom_line(lty="dashed",size=1,alpha=1/3,data = monoculture) +
    geom_line(size=1,data=filter(coculture,condition==x)) +
    geom_errorbar(aes(ymin=y-std,ymax=y+std),size=.5,data = monoculture,alpha=1/3) + 
    geom_errorbar(aes(ymin=y-std,ymax=y+std),size=.5,data = filter(coculture,condition==x)) + 
    geom_point(data = monoculture,alpha=1/3) + 
    geom_point(data=filter(coculture,condition==x))
    
})

# plot enhancement
growth_curves <- c(list(growth_curve_mono), growth_curves_coculture)
growth_curves <- lapply(growth_curves, function(p){
  p + ylab("Log(quantity)") + xlab("Time (h)") +
        scale_y_continuous(limits = c(5,10.8),breaks = 5:10) +
        scale_color_discrete(labels=organsim_fullname,name="Species") +
        theme_bw() +
        theme(legend.position = c(0.7,0.2),
              legend.text = element_text(face = "italic"))
})

# show plot
plot_grid(plotlist = growth_curves,ncol=2,labels = "AUTO")
# ggsave("figure 1.tiff",path="figures")
# export::graph2ppt(file="figures.pptx",append=TRUE)

```

### Comparision of coculture and monoculture quantities

Figure \@ref(fig:monoculture-vs-coculture) shows the variance analysis of the species abundances between monoculture and three cocultures. Results for *E. coli* (A-G) and *P. putida* (H-N) were given by time, as been illustrated on the top subtitle. In x-axis, mono represents the quantity of monoculture, and “1:1000”, “1:1”, “1000:1” represent the quantity of three cocultures. The significance of p-values were showed by “\*” (p<0.05), “\*\*” (p<0.01) or “ns” (p>0.05).


```{r monoculture-vs-coculture, fig.width=12,fig.cap="Analysis the variance of the species abundances between monoculture and three cocultures"}
ref_group <- c("all","none")
library(ggtext)
plots <- lapply(c(1,2), function(i){
  ref <- ref_group[[i]]
  lapply(c(0,0.5,1,2,4,8,24), function(x){
   df <- qPCR_data %>% 
     filter(species == organism[[i]],time==x) %>% 
     select(condition,quantity)
  df$condition <- relevel(df$condition, ref)
  ggplot(df,aes(condition,log10(quantity))) + 
    geom_boxplot() + 
    geom_jitter() + 
    scale_x_discrete(labels = c("mono","1:1000","1:1","1000:1")) +
    scale_y_continuous(expand = c(0,1)) +
    labs(subtitle = paste0("*",organsim_fullname[[organism[[i]]]],"* - ", x, "h")) +
    stat_compare_means(ref.group = ref,label = "p.signif") +
    theme(plot.subtitle = element_markdown()) 
  })
  
})
plots <- unlist(plots,recursive = F)
plot_grid(plotlist = plots,ncol=5,labels = "AUTO")

```


### Ratio (EC/PP) changes in cocultures

Ratios of EC/PP in cocultures were calculated, and ratio changes in cocultures were compared in Figure \@ref(fig:ratio-change).

```{r}
# calculate ratio in cocultures
ratio <- qPCR_data %>% 
  filter(condition %in% c("less","equal","more")) %>% 
  group_by(sample, condition, species, time) %>% 
  mutate(rep = row_number()) %>% 
  ungroup() %>% 
  pivot_wider(names_from = species, values_from = quantity) %>%
  mutate(ratio = EC/PP) %>%
  filter(!is.na(ratio))
```

Figure \@ref(fig:ratio-change) shows the deterministic assembly of *E. coli* and *P. putida* cocultures. (A) the real-time ratio of EC: PP after 0, 0.5, 1, 2, 4, 8 and 24 h cultivation. (B) Analysis of the variances of EC/PP ratios (24h) between cocultures.

```{r ratio-change, fig.width=6, fig.asp=0.618}
ratio.sum <- ratio %>% 
  group_by(sample) %>%
  mutate(y=mean(ratio,na.rm=TRUE),std=sd(ratio,na.rm=TRUE)) %>% 
  select(sample,condition, time, y, std) %>%
  unique()
ratio.sum$condition <- factor(ratio.sum$condition,
                        levels = c("less","equal","more"),
                        labels = c("1:1000","1:1","1000:1")) 
plot_ratio <- ggplot(ratio.sum, aes(time,y,shape=condition,color=condition)) + 
  geom_rect(aes(xmin=23,xmax=25,ymin=0.02,ymax=0.3),
            fill="lightyellow",color="black",alpha=0.1) +
  geom_line(size=1,show.legend = F) +
  geom_point(size=2,show.legend = F) +
  geom_errorbar(aes(ymin=y-std,ymax=y+std),show.legend = F) +
  geom_text(aes(x=9,label=condition),hjust=0,vjust=c(0,0,1),
            data = filter(ratio.sum,time==8),
            show.legend = F) +

  # directlabels::geom_dl(aes(label=condition),method="smart.grid") +
  scale_y_log10(labels=formatC,breaks=10^(-3:3)) +
  labs(x="Time (h)", y="Ratio (EC/PP)") +
  # scale_x_continuous(limits = c(-5,NA)) +
  theme(legend.position = c(0.8,0.75))

ratio_24h <- ratio %>% filter(time==24)
plot_ratio_stats <- ggplot(ratio_24h,aes(condition,ratio,color=condition)) +
  geom_boxplot(fill="lightyellow") + 
  geom_jitter() + 
  stat_compare_means(
    comparisons = list(c("less","equal"),c("less","more"),c("equal","more")),
    label="p.format") +
   scale_x_discrete(breaks=c("less","equal","more"),
                    labels=c("1:1000","1:1","1000:1"))+
  xlab("Condition") + ylab("Ratio (EC/PP)") +
  theme(legend.position = "none",
        panel.background = element_rect(fill="lightyellow"
      ))

plot_grid(plot_ratio,plot_ratio_stats,labels = "AUTO",rel_widths = c(3,2))
# ggsave("figure 2.tiff",path="figures")
# export::graph2ppt(file="figures.pptx",append=TRUE)
```

Notably, the ratio differences were non-significant by time in the logarithmic phase, but were all significant in the stationary phase for every coculture (Figure \@ref(fig:comparison-by-time)). 

For each coculture, significance of variances between each sample were as follows:

- the 1:1000 coculture

```{r}
### stats
ratio1 <- ratio %>% filter(condition=="less")
(p1 <- pairwise.wilcox.test(ratio1$ratio,ratio1$time,p.adjust.method = "BH"))
```

- the 1:1 coculture

```{r}
ratio2 <- ratio %>% filter(condition=="equal")
(p2 <- pairwise.wilcox.test(ratio2$ratio,ratio2$time,p.adjust.method = "BH"))
```

- and the 1000:1 coculture

```{r}
ratio3 <- ratio %>% filter(condition=="more")
(p3 <- pairwise.wilcox.test(ratio3$ratio,ratio3$time,p.adjust.method = "BH"))
```

Figure \@ref(fig:comparison-by-time) shows P-values of pairwise comparison of real-time EC/PP ratios in the “1:1000” (A), “1:1” (B) and “1000:1” (C) cocultures. Circles showed the adjusted p-value for the comparison of the EC/PP ratios between two samples, which were indicated on the top and left (large circle mean large p-value). On this plot, a cross mark was given if the p-value is non-significant (P > 0.05).

```{r comparison-by-time, fig.cap="P-values of pairwise comparison of real-time EC/PP ratios in the “1:1000” (A), “1:1” (B) and “1000:1” (C) cocultures.", fig.asp=0.4}
par(mfrow=c(1,3))
pvalues <- list(p1,p2,p3)
line = 0
cex = 1.2
side = 3
adj=0.15
plots <- lapply(seq_along(pvalues), function(i){
  x <- pvalues[[i]]
  corrplot(x$p.value, 
           type = "lower",
           col = "grey",
           is.corr = FALSE, 
           method = "circle",
           p.mat = x$p.value,
           sig.level = 0.05)
  mtext(LETTERS[[i]], side = side, line = line, cex = cex, adj = adj)

})

```



## Gene expression analysis

# raw data

每个文库测序原始数据量，短序列的比对率。

```{r}
tableS1 <- read.xlsx("./data/tableS1.xlsx")
summary(tableS1)
```

在每个样本中，Aligned reads两个物种的相对比例。

```{r fig.asp=1.5,width=4.5}
ht_counts <- readRDS(file = "./data/rna/ht_counts.rds")
ht_counts$group <- factor(ht_counts$group,levels = c("none_0h","none_4h","none_8h","none_24h","less_0h","less_4h","less_8h","less_24h","equal_0h","equal_4h","equal_8h","equal_24h","more_0h","more_4h","more_8h","more_24h","all_0h","all_4h","all_8h","all_24h"),labels = c("P. putida_0h","P. putida_4h","P. putida_8h","P. putida_24h","1:1000_0h","1:1000_4h","1:1000_8h","1:1000_24h","1:1_0h","1:1_4h","1:1_8h","1:1_24h","1000:1_0h","1000:1_4h","1000:1_8h","1000:1_24h","E. coli_0h","E. coli_4h","E. coli_8h","E. coli_24h"))
ht_counts_total <- ht_counts %>% group_by(sample_id, group, organism) %>%
  summarise(sum_of_reads=sum(count)) %>%
  group_by(sample_id) %>% 
  mutate(proportion=sum_of_reads/sum(sum_of_reads))
samples <- levels(ht_counts_total$group)
plots <- lapply(1:length(samples),function(i){
  sample_group <- samples[[i]]
  df <- filter(ht_counts_total,group==sample_group)
  ggplot(df,aes(x=sample_id, y=proportion,fill=organism)) + 
    geom_bar(stat = "identity",position = "stack") + 
    scale_y_continuous(labels = function(l)paste(format(l*100,digits = 2),"%",sep="")) +
    scale_x_discrete(labels=c("Rep1","Rep2","Rep3")) +
    scale_fill_discrete(name="Organism: ",labels=c("EC"="E. coli","PP"="P. putida")) +
    labs(title = sample_group) +
    theme(legend.text = element_text(face = "italic"),
          legend.position = "none",
          axis.title = element_blank())
})
legend <- get_legend(plots[[1]] + theme(legend.position = "top",legend.direction = "horizontal"))
for (i in c(1,5,9,13,17)){
  plots[[i]] <- plots[[i]] + ylab("Proportion") + theme(axis.text.y = element_text())
}
for (i in 17:20){
  plots[[i]] <- plots[[i]] + xlab("No. of replicates")
}
plot_grid(legend, plot_grid(plotlist = plots,labels = "AUTO",ncol=4),rel_heights = c(1,15),ncol=1)
ggsave("figure S3.tiff",path="figures")
export::graph2ppt(file="figures.pptx",append=TRUE)

# # following is the faceted plot (inconvinent to add subplot labels)
# ggplot(ht_counts_total,aes(x=sample_id, y=proportion,fill=organism)) + 
#   geom_bar(stat = "identity",position = "stack") + 
#   facet_wrap(~group,scale="free_x",ncol=8) +
#   scale_y_continuous(labels = function(l)paste(format(l*100,digits = 2),"%",sep="")) +
#   scale_x_discrete(labels=c("1","2","3")) +
#   scale_fill_discrete(name="Organism: ",labels=c("EC"="E. coli","PP"="P. putida")) +
#   labs(x="No. of replicates",y="Proportion",
#        title=expression(paste("Reads count of ", italic("E. coli")," and ", italic("P. putida"), " in each RNA-seq library"))) +
#   theme(legend.text = element_text(face = "italic"),
#         legend.position = c(0.7,0.1),
#         legend.direction = "horizontal")

```

```{r fig.width=6, fig.asp=0.8}
# only cocultures
plot_grid(legend, plot_grid(plotlist = plots[5:16],labels = "AUTO",ncol=4),rel_heights = c(1,15),ncol=1)
export::graph2ppt(file="figures.pptx",append=TRUE)
```

# Identify gene expression changes

This step is a time consuming step. Use the precalculated DEG if possible.

```{r comparisons}
comparisons.EC <- list(
  c("less_0h","all_0h"),
   c("less_4h","all_4h"),
   c("less_8h","all_8h"),
   c("less_24h","all_24h"),
   c("equal_0h","all_0h"),
   c("equal_4h","all_4h"),
   c("equal_8h","all_8h"),
   c("equal_24h","all_24h"),
   c("more_0h","all_0h"),
   c("more_4h","all_4h"),
   c("more_8h","all_8h"),
   c("more_24h","all_24h"))

comparisons.PP <- list(
   c("less_0h","none_0h"),
   c("less_4h","none_4h"),
   c("less_8h","none_8h"),
   c("less_24h","none_24h"),
   c("equal_0h","none_0h"),
   c("equal_4h","none_4h"),
   c("equal_8h","none_8h"),
   c("equal_24h","none_24h"),
   c("more_0h","none_0h"),
   c("more_4h","none_4h"),
   c("more_8h","none_8h"),
   c("more_24h","none_24h"))
```

```{r run_DESeq2, eval=FALSE}
dds.EC <- readRDS(file = "./data/rna/dds.EC.rds")
dds.PP <- readRDS(file = "./data/rna/dds.PP.rds")

# dds.EC$group <- relevel(dds.EC$group, "all_0h")
# dds.PP$group <- relevel(dds.PP$group, "none_0h")

dds.EC <- DESeq(dds.EC)  # this step is time consuming  标准化
dds.PP <- DESeq(dds.PP)
# dds.EC <- lfcShrink() )  与每个时间点的单菌比较，找差异基因
DEG_results.EC <- myDEG_Results(dds=dds.EC, comparison = comparisons.EC)
DEG_results.PP <- myDEG_Results(dds=dds.PP, comparison = comparisons.PP)
saveRDS(dds.EC,"data/dds.EC.2.rds")
saveRDS(dds.PP,"data/dds.PP.2.rds")
saveRDS(DEG_results.EC,"data/DEG_results.EC.rds")
saveRDS(DEG_results.PP,"data/DEG_results.PP.rds")
```

load precalculated DEG results.

```{r}
dds.EC <- readRDS("data/dds.EC.2.rds")
dds.PP <- readRDS("data/dds.PP.2.rds")
DEG_results.EC <- readRDS("data/DEG_results.EC.rds")
DEG_results.PP <- readRDS("data/DEG_results.PP.rds")
```

# RNA-seq clustering

```{r}
list_of_vsd <- lapply(list(dds.EC,dds.PP),function(dds){
  vst(dds,blind = F)
})
list_of_vsd[[1]]$ratio0 <- factor(list_of_vsd[[1]]$ratio0,
                                  levels = c("none","less","equal","more","all"),
                        labels = c("P. putida","1:1000","1:1","1000:1","E. coli"))
list_of_vsd[[2]]$ratio0 <- factor(list_of_vsd[[2]]$ratio0,
                                  levels = c("none","less","equal","more","all"),
                        labels = c("P. putida","1:1000","1:1","1000:1","E. coli"))

```

```{r fig.width=10,fig.asp=0.618}
list_of_PCA_plot <- lapply(list_of_vsd, function(vsd) {
  myPlotPCA(vsd) + 
    facet_wrap(~time,ncol=4) +
    directlabels::geom_dl(aes(label=ratio0),method = "smart.grid",size=2) +  #文本代替标签 位置标注的不好,改size没用
    scale_color_manual(limits=c("E. coli","1:1000","1:1","1000:1","P. putida"),
                       
                       values = brewer.pal(9,"YlOrRd")[5:9],
                       name="initial ratio\n(EC/PP)") +
    theme(legend.position = "P. putida")
  })

legend <- get_legend(list_of_PCA_plot[[1]] + theme(legend.position = "right"))

plot_grid(plotlist = list_of_PCA_plot,labels = "AUTO",ncol=1)
ggsave("figure 3.tiff",path="figures")
export::graph2ppt(x=plot_grid(plotlist = list_of_PCA_plot,labels = "AUTO",ncol=1),file="PCA.pptx",width=6,height=4)

```

```{r upset plot for EC,fig.width=5}
library(UpSetR)   #可视化
EC_deg <- lapply(DEG_results.EC, function(x) x$gene)
names(EC_deg) <- sapply(comparisons.EC, function(x)x[-2])

upset_deg_EC <- upset(fromList(EC_deg),
                      sets=rev(names(EC_deg)[lapply(EC_deg, length)>0]),
                      nintersects = 10,
                      keep.order = T,
                      mb.ratio = c(0.55, 0.45),
                      order.by = "freq")



```

```{r upset plot for PP,fig.width=5,fig.asp=0.55}
library(UpSetR)

PP_deg <- lapply(DEG_results.PP, function(x) x$gene)
names(PP_deg) <- sapply(comparisons.PP, function(x)x[-2])
pp_selected_sets <- rev(c("less_24h","equal_4h","equal_24h","more_4h","more_8h"))

upset(fromList(PP_deg),
      sets=rev(names(PP_deg)[lapply(PP_deg, length)>0]),
      nintersects = 10,keep.order = T,mb.ratio = c(0.6, 0.4),order.by = "freq")


upset(fromList(PP_deg),
      sets=pp_selected_sets,
      nintersects = 10,keep.order = T,mb.ratio = c(0.6, 0.4),order.by = "freq")
ggsave("figure S8.tiff",path="figures")
```

```{r fig.asp=0.8}
## 差异表达基因个数的变化

# E. coli
EC_up_deg <- lapply(DEG_results.EC, function(x){with(x,gene[expression=="up"])})
EC_dn_deg <- lapply(DEG_results.EC, function(x){with(x,gene[expression=="dn"])})
names(EC_up_deg) <- sapply(comparisons.EC, function(x)x[-2])
names(EC_dn_deg) <- sapply(comparisons.EC, function(x)x[-2])

sapply(EC_up_deg, length) -> up
sapply(EC_dn_deg, length) -> dn
dn <- -dn
deg_count_EC <- rbind(up,dn)
p_deg_count_EC <- reshape2::melt(deg_count_EC) %>% 
  separate(Var2,into = c("ratio0","time")) %>%
  mutate(time=as.numeric(gsub("h","",time))) %>%
  mutate(ratio0=factor(ratio0,levels = c("less","equal","more"),labels = c("1:1000","1:1","1000:1"))) %>%
  ggplot(aes(time,value,group=Var1,color=Var1)) + 
  geom_line(size=1) + 
  # geom_text_repel(aes(label=value)) +
  facet_wrap(~ratio0) + 
  # xlim(c(4,24))+
  scale_color_discrete(name="Gene expression:") +
  scale_y_continuous(labels = function(x){abs(x)}) +
  geom_point() +
  labs(x="Time(h)",y="Count of DEGs",title=expression(paste("DEGs in ",italic("E. coli")))) +
  theme(legend.position = c(0.618,1),
        legend.justification = c(0.5,-0.65),
        legend.direction = "horizontal")


# P. putida
PP_up_deg <- lapply(DEG_results.PP, function(x){with(x,gene[expression=="up"])})
PP_dn_deg <- lapply(DEG_results.PP, function(x){with(x,gene[expression=="dn"])})
names(PP_up_deg) <- sapply(comparisons.PP, function(x)x[-2])
names(PP_dn_deg) <- sapply(comparisons.PP, function(x)x[-2])

sapply(PP_up_deg, length) -> up
sapply(PP_dn_deg, length) -> dn
dn<- - dn
deg_count_PP <- rbind(up,dn)
p_deg_count_PP <- reshape2::melt(deg_count_PP) %>% 
  separate(Var2,into = c("ratio0","time")) %>%
  mutate(time=as.numeric(gsub("h","",time))) %>%
  mutate(ratio0=factor(ratio0,levels = c("less","equal","more"),labels = c("1:1000","1:1","1000:1"))) %>%
  ggplot(aes(time,value,group=Var1,color=Var1)) + 
  geom_line(size=1) + 
  # geom_text_repel(aes(label=value)) +
  facet_wrap(~ratio0) + 
  scale_color_discrete(name="Gene expression:") +
  geom_point() +
  scale_y_continuous(labels = function(x){abs(x)}) +
  # xlim(c(4,24))+
  labs(x="Time(h)",y="Count of DEGs",title=expression(paste("DEGs in ",italic("P. putida")))) +
  theme(legend.position = c(0.618,1),
        legend.justification = c(0.5,-0.65),
        legend.direction = "horizontal")


plot_grid(p_deg_count_EC,p_deg_count_PP,labels = "AUTO",ncol = 1)
ggsave("figure 4.tiff",path="figures")
export::graph2ppt(file="figures.pptx",append=TRUE)
```

```{r}
deg_count_EC %>% abs() %>% sum()
deg_count_PP %>% abs() %>% sum()

```

Venn Diagram

"less"组*E. coli*和"more"组的*P. putida*差异表达基因随时间变化。在这两个组合中，不仅差异表达基因数目多，且其中的很大一部分都是特异性的表达。

```{r }
library(ggVennDiagram)
library(ggplot2)

ratio0 <- c("less","equal","more")
deg_Venn_plot_EC <- lapply(1:3, function(i){
  gene_list <- lapply(DEG_results.EC[(i*4-3):(i*4)], function(x){x$gene})
  ggVennDiagram(gene_list,label = "count",
                category.names = c("0h","4h","8h","24h")) +
    scale_fill_gradient(low="white",high="red",limits=c(0,310)) +
    coord_equal() +
    labs(title=ratio0[[i]]) +
    theme(legend.position = "none",
          plot.title = element_text(hjust=0.5))
})

deg_Venn_plot_PP <- lapply(1:3, function(i){
  gene_list <- lapply(DEG_results.PP[(i*4-3):(i*4)], function(x){x$gene})
  ggVennDiagram(gene_list,label = "count",
                category.names = c("0h","4h","8h","24h")) +
    scale_fill_gradient(low="white",high="red",limits=c(0,310)) +
    coord_equal() +
    labs(title=ratio0[[i]]) +
    theme(legend.position = "none",
          plot.title = element_text(hjust=0.5))
})

plot_grid(plotlist = c(deg_Venn_plot_EC,deg_Venn_plot_PP),
          labels = "AUTO")
ggsave("figure S9.tiff",path="figures")
```

# gseKEGG 分析

GSEA分析需要使用全谱。因此需要重新获得差异表达基因的列表。

```{r}
full_DEG_results.EC <- myDEG_Results(dds.EC,comparison = comparisons.EC,filtered = F)
full_DEG_results.PP <- myDEG_Results(dds.PP,comparison = comparisons.PP,filtered = F)

```

## 计算 GSEA 的结果

### 计算 *E. coli*中 GSEA KEGG 的结果

```{r gseKEGG_results.EC,fig.width=12,fig.asp=1}
library(clusterProfiler)
library(enrichplot)

gseKEGG_results.EC <- lapply(full_DEG_results.EC, function(x){
  if (nrow(x) < 1) return(NULL)
  geneList <- x$log2FoldChange
  names(geneList) <- x$gene
  geneList <- sort(geneList,decreasing = T) #排序
  tryCatch(gseKEGG(geneList, organism = "eco",nPerm = 10000,minGSSize = 30),error=function(e) NULL)
})
```

### 计算 *E. coli* GSEA GO分析的结果

```{r }
library(org.EcK12.eg.db)
full_geneList.EC <- lapply(full_DEG_results.EC, function(x){
  geneList <- x$log2FoldChange
  names(geneList) <- bitr_kegg(x$gene,
                               fromType = "kegg",
                               toType = "ncbi-geneid",
                               organism = "eco")$`ncbi-geneid`
  geneList <- sort(geneList,decreasing = T)
  return(geneList)
})


gseGO_MF_results.EC <- lapply(full_geneList.EC, function(geneList){
  gseGO(geneList, 
        OrgDb = org.EcK12.eg.db,
        ont = "MF")
})

gseGO_CC_results.EC <- lapply(full_geneList.EC, function(geneList){
  gseGO(geneList, 
        OrgDb = org.EcK12.eg.db,
        ont = "CC")
})

gseGO_BP_results.EC <- lapply(full_geneList.EC, function(geneList){
  gseGO(geneList, 
        OrgDb = org.EcK12.eg.db,
        ont = "BP")
})
```

### 计算 *P. putida* 中 GSEA KEGG 的结果

```{r}
gseKEGG_results.PP <- lapply(full_DEG_results.PP, function(x){
  if (nrow(x) < 1) return(NULL)
  geneList <- x$log2FoldChange
  names(geneList) <- x$gene
  geneList <- sort(geneList,decreasing = T)
  tryCatch(gseKEGG(geneList, organism = "ppu",nPerm = 10000,minGSSize = 30),error=function(e) NULL)
})
```

### 计算 *P. putida* 中 GSEA GO 的结果

```{r}
library(AnnotationHub)
ah <- AnnotationHub()
(pp <- query(ah,c("putida","KT2440")))
# pp["AH73771"]
# pp["AH73772"]
# pp["AH73773"]
org.ppu.eg.db <- ah[["AH73772"]]
```

```{r }
full_geneList.PP <- lapply(full_DEG_results.PP, function(x){
  geneList <- x$log2FoldChange
  names(geneList) <- bitr_kegg(x$gene,
                               fromType = "kegg",
                               toType = "ncbi-geneid",
                               organism = "ppu")$`ncbi-geneid`
  geneList <- sort(geneList,decreasing = T)
  return(geneList)
})


gseGO_MF_results.PP <- lapply(full_geneList.PP, function(geneList){
  gseGO(geneList, 
        OrgDb = org.ppu.eg.db,
        ont = "MF")
})

gseGO_CC_results.PP <- lapply(full_geneList.PP, function(geneList){
  gseGO(geneList, 
        OrgDb = org.ppu.eg.db,
        ont = "CC")
})

gseGO_BP_results.PP <- lapply(full_geneList.PP, function(geneList){
  gseGO(geneList, 
        OrgDb = org.ppu.eg.db,
        ont = "BP")
})
```

```{r gseKEGG_plots.EC,fig.width=8,fig.asp=2}
gseKEGG_plots.EC <- lapply(seq_along(gseKEGG_results.EC), function(i){
  x <- gseKEGG_results.EC[[i]]
  label <- names(gseKEGG_results.EC)[[i]]
  if (nrow(data.frame(x)) < 1) return(NULL)
  geneset <- seq(1,nrow(data.frame(x)))
  gseaplot2(x,geneSetID = geneset,pvalue_table = T) +
    labs(title=label) +
    theme(legend.position = "right")
})

for (i in seq_along(gseKEGG_plots.EC)){
  plot <- gseKEGG_plots.EC[[i]]
  label <- names(gseKEGG_results.EC)[[i]]
  if(is.null(plot)){
    gseKEGG_plots.EC[[i]] <- ggdraw(NULL) + draw_figure_label(label, position = "top")
  }
  else{
    gseKEGG_plots.EC[[i]] <- gseKEGG_plots.EC[[i]] + draw_figure_label(label, position = "top")
  }
}

# 只有 less 组（即前3个）有结果
plot_grid(plotlist = gseKEGG_plots.EC[1:4],ncol = 1,labels = "AUTO")

ggsave("figure S10.tiff",path="figures")
# export::graph2ppt(plot_grid(plotlist = gseKEGG_plots.EC[1:4],ncol=1,labels = "AUTO"),file="EC_gseKEGG_gseaplot.pptx")
```

```{r gseGO_analysis,fig.width=12,fig.asp=1}
gseGO_MF_gseplots.EC <- lapply(gseGO_MF_results.EC, function(x){
  if (nrow(data.frame(x)) < 1) return(NULL)
  geneset <- seq(1,nrow(data.frame(x)))
  gseaplot2(x,geneSetID = geneset)
})

# 只有 less 组（即前3个）有结果
plot_grid(plotlist = gseGO_MF_gseplots.EC[1:3])

ggsave("figure S11.tiff",path="figures")


```

## GSEA 结果可视化

### GSEA KEGG dotplot

获取GSEA KEGG结果中的数据框，以便统一绘图。

```{r }
# 计算gsea dotplot需要的数据框
gsea_ck_df <- function(gsea_results = gsea_results){
  gsea_dotplot_df <- vector("list",length = length(gsea_results))
  
  for (i in seq_along(gsea_results)){
    df <- data.frame(gsea_results[[i]])
    if (nrow(df)<1) {
      gsea_dotplot_df[[i]] <- NULL
    }
    else{
      df$comparison <- names(gsea_results[i])
      df$GeneRatio <- (str_count(df$core_enrichment,"/") + 1) / df$setSize
      gsea_dotplot_df[[i]] <- df
    }
    
  }  
  
  return(gsea_dotplot_df)
}
```

```{r }
# df1 是 E. coli gseKEGG 结果的数据框
df1 <- do.call("rbind",gsea_ck_df(gseKEGG_results.EC))

# df2 是 P. putida gseKEGG 结果的数据框
df2 <- do.call("rbind",gsea_ck_df(gseKEGG_results.PP))

```

使用 Kappa 对 GSEA 分析结果的条目进行语义分析和聚类。

```{r}
#' @param df a data frame of gsea result, can be easily obtained by `data.frame(gsea)`
#' @return a kappa matrix
calculate_kappa_matrix <- function(df){
  data <- df[,c("Description","core_enrichment")] %>% 
    separate_rows(.data$core_enrichment,sep="/")
  terms <- data$Description %>% unique() %>% sort()
  genes <- data$core_enrichment %>% unique() 
  nterms <- length(terms)
  ngenes <- length(genes)
  kappa <- matrix(0,nterms,nterms)
  rownames(kappa) <- terms
  colnames(kappa) <- terms
  for (i in 1:nterms){
  	for (j in i:nterms){
  	  if (i==j){
  		kappa[i,j] <- 1
  	  }else{
  		termA <- terms[[i]]
  		termB <- terms[[j]]
  		geneA <- (filter(data,Description==termA))[,"core_enrichment"]
  		geneB <- (filter(data,Description==termB))[,"core_enrichment"]
  		a <- length(intersect(geneA,geneB))   # A,B均有的基因数目
  		b <- length(setdiff(geneB,geneA))     # B有A没有的基因数目
  		c <- length(setdiff(geneA,geneB))     # A有B没有的基因数目
  		d <- length(setdiff(genes,union(geneA,geneB)))  # A,B均没有的基因数目
  		# OA <- (a+d)/ngenes                   # observed agreement
  		# AC <- ((a+c)*(a+b)+(c+d)*(b+d))/ngenes^2  # agreement of chance: is the agreement by chance positive plus the agreement by chance for negative
  		# kappa[i,j] <- (OA-AC)/(1-AC)   # substrate and divide that by 1 minus agreement by chance to standardlize this coefficienct
  		kappa[i,j] <- ((a+d)*ngenes-(a+b)*(a+c)-(c+d)*(b+d))/(ngenes^2-(a+b)*(a+c)-(c+d)*(b+d))
  		kappa[j,i] <- kappa[i,j]
  	  }
  	}
  }
  return(kappa)
}
```

*E. coli* GSEA KEGG 的结果可视化

```{r fig.width=10,fig.asp=1}
kappa <- calculate_kappa_matrix(df1)
tree <- nj(as.dist(1-kappa))
(p <- ggtree(tree) + geom_tiplab() + theme_classic() + xlim(NA,1.5))
term_levels <- (p$data %>% filter(isTip==TRUE) %>% arrange(desc(y)))[,"label"]

data <- df1 %>% separate(comparison,into = c("ratio0","time"),extra = "drop") %>%
  mutate(time=factor(time,levels = c("0h","4h","8h","24h")),
         ratio0=factor(ratio0,levels = c("less","equal","more")),
         Description=factor(Description,levels = term_levels$label))

ggplot(data,aes(time,Description,size=GeneRatio,color=enrichmentScore)) + 
  geom_point() +
  facet_wrap(~ratio0) +
  scale_color_gradient2(low = muted("blue"),high = muted("red"))

```

*P. putida* GSEA KEGG 的结果可视化

```{r fig.width=10}
kappa <- calculate_kappa_matrix(df2)
tree <- nj(as.dist(1-kappa))
(p <- ggtree(tree) + geom_tiplab() + theme_classic() + xlim(NA,1.5))
term_levels <- (p$data %>% filter(isTip==TRUE) %>% arrange(desc(y)))[,"label"]

data <- df2 %>% separate(comparison,into = c("ratio0","time"),extra = "drop") %>%
  mutate(time=factor(time,levels = c("0h","4h","8h","24h")),
         ratio0=factor(ratio0,levels = c("less","equal","more")),
         Description=factor(Description,levels = term_levels$label))

ggplot(data,aes(time,Description,size=GeneRatio,color=enrichmentScore)) + 
  geom_point() +
  facet_wrap(~ratio0) +
  scale_color_gradient2(low = muted("blue"),high = muted("red"))

```

*E. coli* 中全谱GSEA KEGG分析的结果。

```{r,fig.width=8,fig.asp=1}
list_of_gseKEGG_results_df <- lapply(list(df1,df2), function(x){
  if (is.null(x)) return(NULL)
  x %>% 
    separate(comparison, into = c("ratio0","time"),sep="_",extra="drop") %>%
  mutate(time=factor(gsub("h","",time),levels = c(0,4,8,24)),
         ratio0=factor(ratio0, levels = c("less","equal","more")),
         sign=ifelse(enrichmentScore>0,"activated","suppressed"))
})

list_of_gseKEGG_dotplot <- lapply(list_of_gseKEGG_results_df, function(df){
  ggplot(df,aes(time,Description,size=GeneRatio,color=p.adjust)) +
  geom_point() + 
  facet_grid(sign~ratio0,scales = "free_y") +
  scale_color_gradient(limits=c(0,0.05),
                       low="red", high="blue") +
  scale_size_continuous(breaks = c(0.2,0.6,1.0),
                        labels = c("0.2","0.6","1.0"),
                        limits = c(0.2,1.0)) +
    scale_x_discrete(limits=c("0","4","8","24")) +
    labs(x="time(h)",y="") +
    theme(legend.position = "none")
})
```

```{r,fig.width=8,fig.asp=1}
library(grid)
library(gtable)
p1 <- list_of_gseKEGG_dotplot[[1]] + 
  labs(subtitle = expression(paste("Gene set enrichment in ",italic("E. coli")))) +
  theme(legend.position = "right")

# 调整panel的高度
gp <- ggplotGrob(p1)
gtable_show_layout(gp)
facet.rows <- gp$layout$t[grepl("panel",gp$layout$name)]
y.var <- sapply(ggplot_build(p1)$layout$panel_scales_y,
                function(l) length(l$range$range))
gp$heights[facet.rows] <- gp$heights[facet.rows] * y.var
grid.draw(gp)
ggsave("figure S12.tiff",path="figures")


```

*P. putida*中全谱GSEA KEGG分析的结果。

```{r,fig.width=8,fig.asp=1}
library(grid)
library(gtable)
p2 <- list_of_gseKEGG_dotplot[[2]] + 
  labs(subtitle = expression(paste("Gene set enrichment in ",italic("P. putida")))) +
  theme(legend.position = "right")

# 调整panel的高度
gp <- ggplotGrob(p2)
gtable_show_layout(gp)
facet.rows <- gp$layout$t[grepl("panel",gp$layout$name)]
y.var <- sapply(ggplot_build(p1)$layout$panel_scales_y,
                function(l) length(l$range$range))
gp$heights[facet.rows] <- gp$heights[facet.rows] * y.var
grid.draw(gp)
ggsave("figure S13.tiff",path="figures")

```

*E. coli* 中 GSEA GO 的结果

```{r}
d1 <- do.call("rbind",gsea_ck_df(gseGO_BP_results.EC))
d2 <- do.call("rbind",gsea_ck_df(gseGO_CC_results.EC))
d3 <- do.call("rbind",gsea_ck_df(gseGO_MF_results.EC))

plotlist <- lapply(list(d1,d2,d3), function(df){
  df %<>% separate(comparison, into = c("ratio0","time"),sep="_",extra="drop") %>%
  mutate(time=factor(gsub("h","",time),levels = c(0,4,8,24)),
         ratio0=factor(ratio0, levels = c("less","equal","more")),
         sign=ifelse(enrichmentScore>0,"activated","suppressed"))
  ggplot(df,aes(time,Description,size=GeneRatio,color=p.adjust)) +
  geom_point() + 
  facet_grid(sign~ratio0,scales = "free_y") +
  # scale_color_gradient(limits=c(0,0.05),
  #                      low="red", high="blue") +
  # scale_size_continuous(breaks = c(0.2,0.6,1.0),
  #                       labels = c("0.2","0.6","1.0"),
  #                       limits = c(0.2,1.0)) +
    scale_x_discrete(limits=c("0","4","8","24")) +
    labs(x="time(h)",y="") +
    theme(legend.position = "none")
})

```

```{r fig.width=8,fig.asp=2}
plot_grid(plotlist = plotlist,ncol = 1)
ggsave("figure S14.tiff",path="figures")

```

不用全谱的话不科学，用全谱的话没法看。现在应该怎么办？

再三权衡之下，决定对数据进行过滤，得出影响最大的那些通路。

enrichmentScore 是一个重要的参数，先尝试仅显示 enrichmentScore 最大和最小的几个值。

```{r}
list_of_gseKEGG_dotplot <- lapply(list_of_gseKEGG_results_df, function(df){
  x1 <- df %>% arrange(enrichmentScore) %>% head()
  x2 <- df %>% arrange(enrichmentScore) %>% tail()
  data <- rbind(x1,x2)
  ggplot(data,aes(time,Description,size=GeneRatio,color=p.adjust)) +
  geom_point() + 
  facet_grid(sign~ratio0,scales = "free_y") +
  scale_color_gradient(limits=c(0,0.05),
                       low="red", high="blue") +
  scale_size_continuous(breaks = c(0.2,0.6,1.0),
                        labels = c("0.2","0.6","1.0"),
                        limits = c(0.2,1.0)) +
    scale_x_discrete(limits=c("0","4","8","24")) +
    labs(x="time(h)",y="") +
    theme(legend.position = "none")
})

plot_grid(plotlist = list_of_gseKEGG_dotplot,labels = "AUTO",ncol = 1,align = "h")
ggsave("figure S15.tiff",path="figures")

```

这样好像也不合理。

那我们看看几个可用的过滤条件的数值分布情况。

```{r}
p1 <- ggplot(rbind(df1,df2),aes(enrichmentScore)) + geom_histogram(binwidth = 0.1,col="grey") + labs(title = "enrichmentScore")
p2 <- ggplot(rbind(df1,df2),aes(GeneRatio)) + geom_histogram(binwidth = 0.1,col="grey") + labs(title = "GeneRatio")
plot_grid(p1,p2,labels = "AUTO")
```

# 带标记的火山图

```{r}
library(DESeq2)
library(tidyverse)
DEG_results.EC.raw <- myDEG_Results(dds.EC,comparison = comparisons.EC,filtered = FALSE)
DEG_results.PP.raw <- myDEG_Results(dds.PP,comparison = comparisons.PP,filtered = FALSE)


enriched_genes.EC <- lapply(gseKEGG_results.EC,function(x){
  # separate_rows(data.frame(x),"core_enrichment",sep="/")
  df <- data.frame(x)
  if (nrow(df) > 0) {
    separate_rows(df,"core_enrichment",sep="/") %>%
      rename(gene ="core_enrichment")
  }
  else{
    return(NULL)
  }
})
```

```{r fig.width=8,fig.asp=1}
idx <- 1:4
pathways <- unique(do.call("rbind",enriched_genes.EC[idx])$Description)
# color value
color_values <- ggsci::pal_npg("nrc")(length(pathways))
names(color_values) <- pathways
volcano_plots.EC <- lapply(idx, function(i){
  result.raw <- DEG_results.EC.raw[[i]]
  result <- DEG_results.EC[[i]]
  genes <- enriched_genes.EC[[i]]
ggplot() + aes(log2FoldChange,-log10(padj)) +
    geom_point(data = result.raw,color="grey") +
    geom_point(data = result, color="black") +
    geom_point(data=left_join(genes,result),mapping = aes(color=Description)) +
    scale_color_manual(values = color_values) +  
    xlab("Log2Foldchange")+
    theme(legend.position = c(0.01,0.99),
          legend.justification = c(0,1),
          legend.title = element_blank(),
          legend.background = element_blank(),
          legend.box.background = element_rect(color = "grey",fill = NA))
  
})
plot_grid(plotlist = volcano_plots.EC,labels = "AUTO",align = "hv")
ggsave("figure 5.tiff",path="figures")

```

*P. putida*中没有enriched pathway，只能绘制一个黑白配的火山图。

```{r fig.width=8,fig.asp=1}
# PP in "more" coculture, time 0, 4, 8h
idx <- 9:11
volcano_plots.PP <- lapply(idx, function(i){
  result.raw <- DEG_results.PP.raw[[i]]
  result <- DEG_results.PP[[i]]
  ggplot() + aes(log2FoldChange,-log10(padj)) +
    geom_point(data = result.raw,color="grey") +
    geom_point(data = result, color="black") +
     xlab("Log2Foldchange")
})
plot_grid(plotlist = volcano_plots.PP,labels = "AUTO",align = "hv")
ggsave("figure S4.tiff",path="figures")

```

# pathway viewer

```{r}
library(pathview)
pathway_ids.EC <- df1$ID
pathway_geneList_names.EC <- df1$comparison
for (i in seq_along(pathway_ids.EC)){
  idx <- pathway_geneList_names.EC[[i]]
  pathview(gene.data = pathway_ids.EC[[i]],
           pathway.id = pathway_geneList_names.EC[[i]],
           species="eco",
           gene.idtype = "kegg",
           kegg.native = T)

}
```
